---
title: "Final Proj"
author: "Nora Walsh"
date: "2026-02-16"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


Some ideas could be something that has to do with radicals and how that effects elections. 

- Did the Biden administration affect whether people voted for the same party in 2020 than they did in 2024?

Extra Stuff
- outcome variable -> party vote consistency
- explanatory variable -> perceptions toward Biden admin
- control variables -> age, education, income, race 

```{r}
library(tidyverse)
data_2020 <- read_csv("~/Downloads/anes_timeseries_2020_csv_20220210/anes_timeseries_2020_csv_20220210.csv")
data_2024 <- read_csv("~/Downloads/anes_timeseries_2024_csv_20250808/anes_timeseries_2024_csv_20250808.csv")
```
## Milestone 3

Cleaning the data set and making graph with variables of the Biden administration and party voting consistency:

```{r}
# colnames(data_2020)
# colnames(data_2024)
```


Milestone 3

```{r}
merged_data <- full_join(data_2020, data_2024, by = c("V200001" = "V240001"))
merged_data <- merged_data |>
  distinct()

merged_data <- merged_data |>
  filter(V202110x > 0 & V242096x > 0)

merged_data <- merged_data |>
  mutate(
    party_2020 = case_when(
      V202110x == 1 ~ "Democrat",
      V202110x == 2 ~ "Republican",
      TRUE ~ "Other"),
    party_2024 = case_when(
      V242096x == 1 ~ "Democrat",
      V242096x == 2 ~ "Republican",
      TRUE ~ "Other"))

merged_data <- merged_data |>
  mutate(party_switch = case_when(
    party_2020 == party_2024 ~ "Same Party",
    party_2020 != party_2024 ~ "Switched Party",
    TRUE ~ NA_character_
  ))

# glimpse(merged_data)
table(merged_data$party_switch)

write.csv(merged_data, "merged_cleaned_data.csv", row.names = FALSE)
```


```{r}
merged_data <- merged_data |>
  filter(!is.na(V241200), V241200 > 0) |>
  mutate(
    biden_approval = case_when(
      V241200 %in% 1:2 ~ "Approve",
      V241200 %in% 4:5 ~ "Disapprove",
      V241200 == 3 ~ "Neutral"),
    switched = party_2020 != party_2024)

ggplot(merged_data, aes(x = biden_approval, fill = switched)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("TRUE" = "tomato", "FALSE" = "steelblue"),
                    labels = c("Stayed", "Switched")) +
  labs(title = "Party Switching by Biden Approval",
       x = "Biden Job Approval",
       y = "Proportion of Respondents",
       fill = "Party Switch?") +
  theme_minimal()
```

This shows whether or not people who disapproved of Biden were likely to switch parties in 2024 compared to those who approved of Biden. I would expect to see a higher level of red which means switch over the disapproval but it is actually equal for approval and disapproval. The neutral party is even between stayed and switched. 

Milestone 4
```{r}

```

```{r}
ggplot(merged_data, aes(x = as.numeric(factor(biden_approval)), y = as.numeric(switched))) +
  geom_jitter(alpha = 0.3, height = 0.05, color = "steelblue") +
  geom_smooth(method = "lm", color = "tomato", se = TRUE) +
  scale_x_continuous(breaks = 1:3, labels = c("Approve", "Neutral", "Disapprove")) +
  labs(title = "Biden Approval and Party Switching",
       x = "Biden Approval",
       y = "Switched Party (0 = No, 1 = Yes)") +
  theme_minimal()
```

This shows the opposite of my hypothesis which is that people who approved Biden were slightly more likely to switch parties as there is a slight downward slop. Neutral still falls int he middle. Since the grey is so large there is lots of uncertainty so the trend is not statistically significant. 




