---
title: "Approval Rates Affecting Party Switches"
author: "Nora Walsh"
date: "2026-04-29"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction

Did approval rates of the Biden administration affect whether people voted republican or democrat the next year?

Hypothesis
People who approved of the Biden administration voted for democrats the next year. 

This is important as it can show if an administration's actions can have true affect on future elections. If a democratic party has poor execution while in office, people may switch and vote republicans. This same concept applies vice versa. Specifically looking into the Biden administration and the approval rating, can give insight into whether or not approval affects party switching. This can show people who campaign how important execution and being liked is to reelection. It will also demonstrate the success of an administration and if that gets more people to stay. Furthering the point of how execution of an administration can determine if a party stays in office for longer. In the end, this could change who parties have run for them if it means they could be risking future party reelection. 


```{r, echo=FALSE}
library(tidyverse)
data_2020 <- read_csv("~/Downloads/anes_timeseries_2020_csv_20220210/anes_timeseries_2020_csv_20220210.csv")
data_2024 <- read_csv("~/Downloads/anes_timeseries_2024_csv_20250808/anes_timeseries_2024_csv_20250808.csv")
```
## Data Section


```{r, echo=FALSE}
# colnames(data_2020)
# colnames(data_2024)
```

The table shows that where democrats voted the year after the Biden administration. To break it down, 858 democrats stayed while 15 voted other and 21 voted republican meaning 21 switched. The overall party switch count is 873 stayed (FALSE) while 21 switched (TRUE). The third table shows that 13.4% switched from the disapproved. So only about 13 out of 100 people who did not approve of Biden switched. This shows little impact of poor performance but there is still an impact on switching since there was little switch. For those who approve of Biden, about 4 out of 590 switched which is a .7% showing what was predicted and that almost all who approved would stay. It is 3 tables showing similar data confirming that those who disapprove were more likely to switch. The data supports the hypothesis as the disapprovers switch at a higher rate. 

```{r, include=FALSE}
merged_data <- full_join(data_2020, data_2024, by = c("V200001" = "V200001")) |>
  distinct() |>
  filter(V202110x > 0 & V242096x > 0) |>
  mutate(
    party_2020 = case_when(
      V202110x == 1 ~ "Democrat",
      V202110x == 2 ~ "Republican",
      TRUE ~ "Other"),
    party_2024 = case_when(
      V242096x == 1 ~ "Democrat",
      V242096x == 2 ~ "Republican",
      TRUE ~ "Other"))

merged_data <- merged_data |>
  mutate(party_switch = case_when(
    party_2020 == party_2024 ~ "Same Party",
    party_2020 != party_2024 ~ "Switched Party",
    TRUE ~ NA_character_
  ))

dem_to_rep <- merged_data |>
  filter(party_2020 == "Democrat", party_2024 == "Republican")

table(merged_data$party_switch)
write.csv(dem_to_rep, "merged_cleaned_data.csv", row.names = FALSE)
```


```{r, echo=FALSE}
dem_to_rep <- dem_to_rep |>
  filter(!is.na(V241200), V241200 > 0) |>
  mutate(
    biden_approval = case_when(
      V241200 %in% 1:2 ~ "Approve",
      V241200 %in% 4:5 ~ "Disapprove",
      V241200 == 3 ~ "Neutral"),
    switched = party_2020 != party_2024)
dem_only <- merged_data |>
  filter(party_2020 == "Democrat") |>
  filter(!is.na(V241200), V241200 > 0) |>
  mutate(
    biden_approval = case_when(
      V241200 %in% 1:2 ~ "Approve",
      V241200 %in% 4:5 ~ "Disapprove",
      V241200 == 3 ~ "Neutral"),
    switched = party_2024 == "Republican")

table(dem_only$party_2020, dem_only$party_2024)
table(dem_only$switched)
table(dem_only$biden_approval, dem_only$switched)
```
This graph shows the outcome variable or dependent variable which is just showing who stayed democrat and who switched to republican. Party switching is the variable. The independent will be shown in a later graph which is approval, disapproval, and neutral. A very high percentage of democrats did not switch parties and this helps predict that those who disapprove are more likely to switch. This is already showing a pattern with the hypothesis since so many democrats stayed. 

```{r, echo=FALSE}
library(ggplot2)
ggplot(dem_only, aes(x = switched)) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("FALSE" = "Stayed Democrat", "TRUE" = "Switched to Republican")) +
  labs(
    title = "Party Switching Among Democratic Voters",
    x = "Party Switch",
    y = "Number of Respondents"
  ) +
  theme_minimal()
```


## Results Section


The bar chart shows how many people switched based on if they approved, disapproved, or had a neutral opinion on Biden. The chart shows that those who disapproved of Biden were more likely to switch and vote republican the next election by a much higher rate than those who approved. It is much taller than both neutral and approved and nuetral and approve were almost equal showing they switched at a much lower rate. It can not claim direct causation because it is cross sectional, observational study so that means correlation rather than causation. 



```{r, echo=FALSE}
ggplot(dem_to_rep, aes(x = biden_approval, fill = switched)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "tomato", "FALSE" = "steelblue"),
                    labels = c("FALSE" = "Stayed Democrat", "TRUE" = "Switched")) +
  labs(title = "Democrat to Republican Switching by Biden Approval",
       x = "Biden Job Approval",
       y = "Proportion of Respondents",
       fill = "Party Switch?") +
  theme_minimal()
```






The regression shows that democrats who disapproved of Biden were about 12.7% points more likely to switch to republican in the next election than those who approved. This percent shows statistical significance and confirms the hypothesis. The p value is 7e-15 which is very statistically significant. The R squared value is .066 which means Biden's approval explains about 6.6% of the party switching which though small, still matters and is significant. The data supports the hypothesis that the people who approve would continue to vote democrat. 

```{r, echo=FALSE}
ggplot(dem_only, aes(x = as.numeric(factor(biden_approval)), y = as.numeric(switched))) +
  geom_jitter(alpha = 0.3, height = 0.05, color = "steelblue") +
  geom_smooth(method = "lm", color = "tomato", se = TRUE) +
  scale_x_continuous(breaks = 1:3, labels = c("Approve", "Neutral", "Disapprove")) +
  labs(title = "Democrat to Republican Switching by Biden Approval",
       x = "Biden Approval",
       y = "Switched Party (0 = No, 1 = Yes)") +
  theme_minimal()
```


```{r, echo=FALSE}
model <- lm(switched ~ biden_approval,
             data = dem_only)
summary(model)
library(knitr)
table(dem_only$biden_approval, dem_only$switched) |>
  kable()
```


## Conclusion

This project looked into whether the approval of the Biden administration had influence on Democratic voters to switch Republican in the next election. The hypothesis predicted that those who approved would continue to vote democrat while those who disapproved would switch. The data supported the hypothesis and the data suggests that approval ratings do have an effect on whether voters will party switch. There are other variables bot accounted for in the data such as economics, ideals on social issues, and certain candidate preferences. To improve the project, there would need to be more variables and perhaps a longer sample. The project demonstrates how voters feel about the administration can affect which party they are going to vote for in the future. This shows the importance of public approval for a party long tern success.  

## AI Usage Statement 

AI was used to support me in the project when I had to edit some of my data to make the project flow more well. It helped to clean up some of my data as when I first started I had a lot. Overall, there was minimum AI usage except for my coding when I needed guidance because I was stuck or had to redo some things. 

