---
title: "Final Proj"
author: "Nora Walsh"
date: "2026-02-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


Some ideas could be something that has to do with radicals and how that effects elections. 

- Did the Biden administration affect whether people voted for the same party in 2020 than they did in 2024?

Extra Stuff
- outcome variable -> party vote consistency
- explanatory variable -> perceptions toward Biden admin
- control variables -> age, education, income, race 

```{r}
library(tidyverse)
data_2020 <- read_csv("~/Downloads/anes_timeseries_2020_csv_20220210/anes_timeseries_2020_csv_20220210.csv")
data_2024 <- read_csv("~/Downloads/anes_timeseries_2024_csv_20250808/anes_timeseries_2024_csv_20250808.csv")
head(firstdata)
colnames(firstdata)
```
## Milestone 3

Cleaning the data set and making graph with variables of the Biden administration and party voting consistency:

```{r}
colnames(data_2020)
colnames(data_2024)
```


```{r}
merged_data <- full_join(data_2020, data_2024, by = c("V200001" = "V240001"))
merged_data <- merged_data |>
  distinct()

merged_data <- merged_data |>
  filter(V202110x > 0 & V242096x > 0)

merged_data <- merged_data |>
  mutate(
    party_2020 = case_when(
      V202110x == 1 ~ "Democrat",
      V202110x == 2 ~ "Republican",
      TRUE ~ "Other"),
    party_2024 = case_when(
      V242096x == 1 ~ "Democrat",
      V242096x == 2 ~ "Republican",
      TRUE ~ "Other"))

merged_data <- merged_data |>
  mutate(party_switch = case_when(
    party_2020 == party_2024 ~ "Same Party",
    party_2020 == party_2024 ~ "Switched Party",
    TRUE ~ NA_character_
  ))

glimpse(merged_data)
table(merged_data$party_switch)

write.csv(merged_data, "merged_cleaned_data.csv", row.names = FALSE)
```

ggplot(merged_data, aes(x = party_2020, fill = party_2024)) + 
geom_bar(position = "dodge") +
labs(title = "2020 vs 2024 Party Vote" 
)
```{r}
ggplot(merged_data, aes(x = party_2020, fill = party_2024)) + 
  geom_bar(position = "dodge") +
  labs(title = "2020 vs 2024 Party Vote",
       x = "2020 Vote",
       y = "Number of Respondents",
       fill = "2024 Vote") +
  scale_fill_manual(values = c("Democrat" = "steelblue",
                               "Republican" = "indianred",
                               "Other" = "gray")) +
  theme_minimal()c
```


